Online Multiclass Linear Classification with Bandit Feedback

### 2019-01-16 arxiv-version/np-hardness-of-the-weak-labeling-problem.tex
add np-hardness proof

Formally, the input to the decision problem are positive integers $d, T, K$ and
a sequence $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in \{0,1\}^d \times
\{1,2,\dots,K, \overline{1}, \overline{2}, \dots, \overline{K}\}$ of strongly
and weakly labeled examples. The numbers $1,2,\dots,K$ correspond to strong
labels and the symbols $\overline{1}, \overline{2}, \dots, \overline{K}$
correspond to weak labels. We adopt the convention that $\overline{\overline{i}}
= i$ for any positive integer $i$. The goal is to determine whether or not there
exist vectors $w_1, w_2, \dots, w_K \in \R^d$ that satisfy the following
conditions for all $t=1,2,\dots,T$,
\begin{align*}
y_t \in \{1,2,\dots,K\} \qquad & \Longrightarrow \qquad & \forall i \in \{1,2,\dots,K\} \setminus \{y_t\} \qquad \ip{w_{y_t}}{x_t} & > \ip{w_i}{x_t} \; , \\
y_t \in \{\overline{1}, \overline{2},\dots, \overline{K}\} \qquad & \Longrightarrow & \qquad \exists i \in \{1,2,\dots,K\} \qquad \ip{w_i}{x_t} & > \ip{w_{\overline{y_t}}}{x_t} \; .
\end{align*}

TODO

>>>>

Formally, the weak labeling problem can be described as below: 
\begin{figure}[H]
\begin{framed}
\begin{center}
    \textbf{Weak Labeling}
\end{center}
\textbf{Given: } A set of feature-label pairs: $\{(x_t, y_t)\}_{t=1}^T \subseteq \{0,1\}^d \times \{1,2,\ldots, K, \overline{1}, \overline{2}, \ldots, \overline{K}\}$. \\
\textbf{Question: } Do there exist $ w_1, \ldots, w_K \in \R^d$ such that for all $t=1, \ldots, T$,  
\begin{align*}
y_t \in \{1,2,\dots,K\} \qquad & \Longrightarrow \qquad & \forall i \in \{1,2,\dots,K\} \setminus \{y_t\} \qquad & \ip{w_{y_t}}{x_t}  > \ip{w_i}{x_t} \; , \\
y_t \in \{\overline{1}, \overline{2},\dots, \overline{K}\} \qquad & \Longrightarrow & \qquad \exists i \in \{1,2,\dots,K\} \qquad & \ip{w_i}{x_t} > \ip{w_{\overline{y_t}}}{x_t} \; .
\end{align*}
\end{framed}
\end{figure}

%Formally, the input to the decision problem are positive integers $d, T, K$ and
%a sequence $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T) \in \{0,1\}^d \times
%\{1,2,\dots,K, \overline{1}, \overline{2}, \dots, \overline{K}\}$ of strongly
%and weakly labeled examples. The numbers $1,2,\dots,K$ correspond to strong
%labels and the symbols $\overline{1}, \overline{2}, \dots, \overline{K}$
%correspond to weak labels. We adopt the convention that $\overline{\overline{i}}
%= i$ for any positive integer $i$. The goal is to determine whether or not there
%exist vectors $w_1, w_2, \dots, w_K \in \R^d$ that satisfy the following
%conditions for all $t=1,2,\dots,T$,
%\begin{align*}
%y_t \in \{1,2,\dots,K\} \qquad & \Longrightarrow \qquad & \forall i \in \{1,2,\dots,K\} \setminus \{y_t\} \qquad \ip{w_{y_t}}{x_t} & > \ip{w_i}{x_t} \; , \\
%y_t \in \{\overline{1}, \overline{2},\dots, \overline{K}\} \qquad & \Longrightarrow & \qquad \exists i \in \{1,2,\dots,K\} \qquad \ip{w_i}{x_t} & > \ip{w_{\overline{y_t}}}{x_t} \; .
%\end{align*}

The hardness proof is based on a reduction from the set splitting problem, which is proven to be NP-complete by Lovasz \chenyu{Reference needed}, to our weak labeling problem. The reduction is adapted from Avrim Blum \chenyu{Reference needed}. 
\begin{figure}[h]
\begin{framed}
\begin{center}
    \textbf{Set Splitting}
\end{center}
\textbf{Given: } A finite set $S$ and a collection $C$ of subsets $c_i$ of $S$. \\
\textbf{Question: } Do there exist disjoint sets $S_1$ and $S_2$ such that $S_1 \cup S_2 = S$ and $\forall i, c_i\not\subseteq S_1$ or $c_i\not\subseteq S_2$? 
\end{framed}
\end{figure}

Below we show the reduction. Suppose we are given a set splitting problem with inputs: 
\begin{align*}
    &S = \{1, 2, \ldots, N\}, \\
    &C = \{c_1, \ldots, c_M\}, \\ 
    &\forall i, c_i \subseteq S. 
\end{align*}

Then we create the weak labeling instance as follows. Let $d=N+1$ and $K=3$. Define $\zero$ as the zero vector $(0,\ldots, 0)\in \R^N$ and $\e_i$ as the $i$-th standard vector $(0,\ldots, 1, \ldots, 0)\in \R^N$). Then we include all the following feature-label pairs: 
\begin{itemize}
    \item Type 1: $(x,y)=((\zero,1), 3)$,
    \item Type 2: $(x,y)=((\e_i,1), \overline{3})$, $\forall i\in [N]$,
    \item Type 3: $(x,y)=\left(\left(\sum_{i\in c_j} \e_i, 1\right), 3\right)$, $\forall j\in[M]$.  
\end{itemize}

For example, if we have $S=\{1,2,3\}$, $C=\{c_1, c_2\}$, $c_1 = \{1,2\}$, $c_2=\{2,3\}$, then we create the weak labeling sample set as: 
\begin{align*}
    \{((0,0,0,1),3), ((1,0,0,1),\overline{3}), ((0,1,0,1),\overline{3}), ((0,0,1,1),\overline{3}), ((1,1,0,1),3), ((0,1,1,1),3)\}.
\end{align*}
The following lemma shows that answering this weak labeling problem is equivalent to answering the original set splitting problem. 
\begin{lemma}
Suppose we are given a set splitting problem, and we construct a weak labeling problem as described above. Then the set splitting problem has a solution $\Longleftrightarrow$ the constructed weak labeling problem has a solution. 
\end{lemma}
\begin{proof}
 $(\Longrightarrow)$ Let $S_1, S_2$ be the solution of the set splitting problem. Define
 \begin{align*}
     w_1 = \left(a_1, a_2, \cdots, a_N, -\frac{1}{2}\right), 
 \end{align*}
 where $\forall i\in[N]$, $a_i=1$ if $i\in S_1$ and $a_i=-N$ if $i\notin S_1$. 
 Similarly, define
 \begin{align*}
     w_2 = \left(b_1, b_2, \cdots, b_N, -\frac{1}{2}\right), 
 \end{align*}
 where $\forall i\in[N]$, $b_i=1$ if $i\in S_2$ and $b_i=-N$ if $i\notin S_2$. 
 Finally, define
 \begin{align*}
     w_3 = (0,0,\cdots, 0), 
 \end{align*}
 the zero vector. To see this is a solution for the weak labeling problem, we verify separately for Type 1-3 samples defined above. For Type 1 sample, we have
 \begin{align*}
     \ip{w_3}{x} = 0 > -\frac{1}{2} = \ip{w_1}{x}=\ip{w_2}{x}. 
 \end{align*}
 For a Type 2 sample that corresponds to index $i$, we have either $i\in S_1$ or $i\in S_2$ because $S_1\cup S_2 = [N]$ is guaranteed. Thus, either $a_i=1$ or $b_i=1$. If $a_i=1$ is the case, then
 \begin{align*}
     \ip{w_1}{x}=a_i-\frac{1}{2}=\frac{1}{2} > 0 = \ip{w_3}{x}; 
 \end{align*}
 similarly if $b_i=1$, we have $\ip{w_2}{x}>\ip{w_3}{x}$. \\
 For a Type 3 sample that corresponds to index $j$, Since $c_j \not\subset S_1$, there exists some $i'\in c_j$ and $i'\notin S_1$. Thus we have $x_{i'}=1$, $a_{i'}=-N$, and therefore
 \begin{align*}
     \ip{w_1}{x}= a_{i'}x_{i'} + \sum_{i\in [N]\backslash\{i'\}} a_ix_i - \frac{1}{2} \leq -N + (N-1)-\frac{1}{2} < 0 = \ip{w_3}{x}. 
 \end{align*}
 Because $c_j \not\subset S_2$ also holds, we also have $\ip{w_2}{x}<\ip{w_3}{x}$. This direction is therefore proved. \\
 \ \\
 $(\Longleftarrow)$ Given the solution $w_1, w_2, w_3$ of the weak labeling problem, we define 
 \begin{align*}
     S_1 = \left\{i\in [N] \Big| \ip{w_1-w_3}{(\e_i, 1)} > 0 \right\}, \\
     S_2 = \left\{i\in [N] \Big| \ip{w_2-w_3}{(\e_i, 1)} > 0 \right\}.
 \end{align*}
 It is not hard to see $S_1\cup S_2 = [N]$. Note that $(\e_i, 1)$ is the feature vector for Type 2 samples. Because Type 2 samples all have label $\overline{3}$, for any $i\in [N]$, one of the following must hold: $\ip{w_1-w_3}{(\e_i, 1)}>0$ or $\ip{w_2-w_3}{(\e_i, 1)}>0$. Thus $i\in S_1$ or $i\in S_2$.

 Now we show $\forall j$, $c_j \not\subset S_1$ and $c_j \not\subset S_2$ by contraction. Without loss of generality, suppose there exists some $j$ such that $c_j \subset S_1$. By our definition of $S_1$, we have $\ip{w_1-w_3}{(\e_i, 1)} > 0$ for all $i\in c_j$. Therefore, 
 \begin{align*}
     \sum_{i\in c_j}\ip{w_1-w_3}{\left(\e_i, 1\right)} = \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, |c_j|\right)} > 0. 
 \end{align*}
 Because Type 1 sample has label $3$, we also have
 \begin{align*}
     \ip{w_1-w_3}{\left(\zero, 1\right)} < 0. 
 \end{align*}
 Combining the above two inequalities, we get
 \begin{align*}
     \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, 1\right)} = \ip{w_1-w_3}{\left(\sum_{i\in c_j}\e_i, |c_j|\right)} - (|c_j|-1)\ip{w_1-w_3}{\left(\zero, 1\right)}>0.  
 \end{align*}
 Note that $\left(\sum_{i\in c_j}\e_i, 1\right)$ is a feature vector for Type 3 samples. Thus the above inequality contradicts that Type 3 samples have label 3. Therefore, $c_j \not\subset S_1$. Similarly we can prove $c_j \not\subset S_2$. 
\end{proof}
